% \chapter*{\begin{center}\Large{Acknowledgements --- It Takes a Village}\end{center}}
\chapter*{\begin{center}\Large{Acknowledgements}\end{center}}
{
\fontsize{10}{12}\selectfont

% Regrettably, I haven't had the time to write an acknowledgements section. I'm
% looking forward to adding this as soon as I get chance. I think the
% acknowledgement section of a thesis is a really important thing: it's a place
% where the candidate gets to lay a personal touch an academic document like this
% can have (the only one that's really permitted). The fact that the only personal
% part of a thesis is about \emph{other} people is telling. A PhD isn't something
% anybody does in isolation, and it's important to be aware of how much we owe to
% the folk we're surrounded by.
% 
% I certainly couldn't have produced this without a mammoth amount of help and
% guidance from all sorts of people, often spanning years and years. My name's on
% the paper in some ceremony after all of this is over, but the work they've put
% in has been essential to delivering a complete thesis just the same. If this is
% the only recognition they'll get, isn't it important to shout about their
% support where I can? I really care about paying deference to all those people.
% It takes a village to raise a child because of the time, support, and community
% a baby needs --- that's just as true for a PhD student, and it'd be such a shame
% for all those folk not to get their day in the sun too.
% 
% So, I regret that this still isn't written. With any luck I'll revisit this
% before I submit, but I doubt it. I'll take the time \emph{now} to make the
% content worth examining, and take the time \emph{later} to write up these
% acknowledgements while I wait for my viva.

{
    \centering
    \emph{This one really did take a village.}
}

First: I'm extremely grateful to Tim Storer, my supervisor. You've always been
an understanding and supportive advisor — I'm sure I didn't always make that
easy! Thank you for your patience and your guidance.

A big thank you, too, to International Data Flows Ltd, who sponsored the work in
this thesis. Particular thanks to Paul Wallis and Fergus Cloughley, who
supported me throughout with thoughtful encouragement and opportunities to grow.

My partner, Marta, has been extremely supportive at every turn. You've listened
countless times to rambles about experiments, about ideas and papers and
opportunities, about frustrations and limitations, and about chapters and
figures and tables --- and now corrections, too. Who knew listening could be an
endurance sport? Countless times when I've needed a boost, you've been there to
cheer me on and keep me above water. Having you with me though the PhD ---
especially these last few months of writing! --- has felt like having a copilot
helping keep this plane in the sky. I'm very grateful. Thank you thank you
\emph{thank you}.

My parents, Paul and Liz, have waited a long while to celebrate this moment.
Guess what? It's finally time! Thank you for cheering me on through the years
and reminding me when to keep pushing, and when to stop and breathe. When people
ask me about my parents, I always tell them how lucky I am to have had you
two\ldots{} here I go again. As I'm writing this, you're getting ready to come
to London and celebrate getting this PhD over the line. I can't wait.

Thanks to Jessica Enright, who graciously agreed to be a second supervisor
late in the project and offered advice, encouragement, and support every time I
asked. Thanks also to Rob Dekkers for taking the helm in the early stages of
this thesis.

I'm also grateful to Phil Trinder and Steve Marsh for an engaged, lively, and
surprisingly enjoyable viva — thanks also to Steve Brewster for chairing it.

Ben Bumpus, the idea of getting through this viva without our morning coffees,
wild enthusiasm, climbing sessions and the opportunities to catch up between
them, and jaunts to Nigeria and Rwanda to teach and adventure is unthinkable.
Would all these years have been the same without macchiatos in Little Italy and
sparring on any and every discussion topic we could find? I doubt it! You're a
terrific friend. Thanks for your inexhaustable joy and enthusiasm, and for being
a rock-steady ally, no questions asked.

William Kavanagh, I'm still taken aback that we didn't just get RPGLite
released, but actually got folk to play. If I realised when we started
collaborating that I was signing up to all those months beta testing, late
nights designing configurations of the game, and panicked moments after release
making sure servers were stable and intact\ldots{}I reckon I'd have jumped at it
even quicker. What a highlight! Thanks for your tireless friendship.

There are lots of people from Glasgow University whose friendship, support,
and/or wisdom and guidance have helped get me to this point.
Marco Cook always had a joke up his sleeve, a cheeky grin to pair it with, and
the best eagle impression in the business. Craig Reilly was the best DM a player
could ask for, a perennial friend, and schooled me on food, coffee, and all
things dweeby. Matt Barr was an absolute rockstar: he seemed to hold the GA
programme up like some kind of academic Atlas, and somehow still found the time
(and energy!) to make it a lovely place to work; the apprentices and the team
are lucky to work with you, and so was I! Likewise, everybody working on the GA
programme --- particularly Waqar Syed Nabi and Derek Sommerville --- were a
delight, and I had a lovely time building that proverbial plane together while
we were flying it. The graduate apprentices themselves deserve a mention too,
for being such a joy to teach; you were all absolutely wonderful, but special
shout-outs should go to Daniel Barnett, Goudham Souresh, Jade Nelson, Kathleen
West, Leia Matchett, Lixin Su, and Robert Ranson. Quintin Cutts' wise words were
always right on the money. Lovisa Sundin was always smart as a whip and twice as
funny, and our coffees and catch-ups were the most reliable tonic. Gözel
Shakeri's indominable enthusiasm was infectious, and lit up every room she was
in. Patrizia Di Campli San Vito had a listening ear for anybody and everybody,
and I don't know how any of us would have kept going some days without her
constant support and happy smile. Frances Cooper was a force of nature that
seemed to bring people together and bring the best out of all of us, day in, day
out. Fionnuala Johnson had the most encouraging and tenacious soul, and teaching
in Nigeria and Rwanda together was unforgettable. David Maxwell's positive,
joyful spirit was an inspiration --- oh, and David! Your beautiful and
comprehensive acknowledgements section inspired my own! Ting Su was caring,
kind, and always checked in to make sure things were going OK. Joe Armstrong
weathered all our storms with us and kept the department running despite all
odds.
I think it's easy to underestimate how important the community around you is
when you're taking on something like a PhD, and the little community in Lilybank
Gardens was the best anybody could ask for.

I had lots of help getting the thesis over the line in time --- thanks to for
the hours reading the thesis and providing commentary to Derek Leung, who truly
went beyond the call of duty, and to Laurence Tratt for reading through
extremely rough chapter drafts and leaving extremely constructive, helpful
feedback on them. Your feedback was invaluable! Thanks also to Marta and my Mum
and Dad who spent, I'm sure, more evenings over their festive holidays than they
ideally would have to proofreading chapters ``one last time'' quite a few times.
When the viva loomed, Milad Kazemi --- who must have been one of the most
cheerful and friendly faces in all of London and the most fascinating repository
of knowledge --- provided sage advice and a calming counterbalance to pre-viva
nerves, and set up a mock viva without skipping a beat like a world champion of
doing good deeds. Thank you!

Thanks to Catherine Wallis and Jamie Wallace for pro-tips when we moved to the
big city, for your enthusiasm for RPGLite, and for always cheering me on. I
can't wait to celebrate your lives together --- just a few weeks away as I'm
writing these acknowledgements! Time flies! Thanks also to Benji, for being a
very good boy.

Thanks, also, to my Gran, Grandma, and Grandad, whose excitement at hearing I'd
passed my viva had clearly been brewing for seven years. I'm sorry you had to
wait so long for that good news! Thank you for being so proud when it finally
happened, and for all your love and kindness in the meantime.

Lots of friends listened to stories about writing this thesis over my weekends
week in, week out. I'm lucky to have friends who were so patient and
understanding when the pressures of writing were getting the best of me. Thanks
to George Campbell for being the friendliest face in Putney and the best coffee
our side of the Thames. Big thanks to Andrea Fernandez-Buitrago, Boyang Pang,
Chloe Taylor, Derek Leung, Ilia Teterin, Muyiwa Iyowu, Nejla Lowe, Rick
Newton-Rogers, Sevki Harasci, Xia Liu, and Zoltân Szkeres for keeping me sane
sprint after sprint in my sojourn into industry. Thanks to David Creigh for all
the lunchtime catch-ups, walks around Glasgow, and listening to my mad ideas
throughout a decade of friendship. Thanks also to the coolest cats in Glasgow:
Alexander Morgan, Emma Toussaint, Laura Morgan, Leisha Hussein and Rita
Toussaint. Cass Brooks brought her curious, clever, and funny A-game every time
we caught up, without fail, from so many miles away. Zak Bevan's patience for my
lack of calendar management deserves a thanks all of its own (as with Cass and
countless others!) --- thanks too, Zak, for brilliant conversation, top-notch
board game recommendations, and more evenings watching classic sci-fi than I can
count. Thanks to Mattia Salnitri for advice and collaboration with some of the
ideas in the early stages of this PhD. Thanks to Matilde Araldo and Piero Araldo
for being wonderful hosts every visit and rewarding my broken Italian with
focaccia so good it should be illegal. Thanks to Connor Cloughley, Anna-Sofia
Myszka, and Pavlos Evangelidis at Obashi, who always made visits to the office
something to look forward to. Thanks to Francesco Porro and to Luca Amato for
your enthusiasm for, and feedback on, RPGLite, which was a delightful surprise.



% - [X] Tim Storer
% - [X] Jessica Enright
% - [X] Steve Marsh
% - [X] Phil Trinder
% - [X] Steve Brewster
% - [X] Marta Araldo
% - [X] Liz Wallis (Mum)
% - [X] Paul Wallis (Dad)
% - [X] Paul Wallis
% - [X] Fergus Cloughley (sp?)
% - [X] Obashi Technologies
% - [X] Laurence Tratt
% - [X] Derek Leung
% - [X] Benjamin Merlin Bumpus
% - [X] Andrea Fernandez-Buitrago
% - [X] Nejla Lowe
% - [X] Chloe Taylor
% - [X] Zoltân Szkeres (sp?)
% - [X] Muyiwa Iyowu
% - [X] Ilia Teterin
% - [X] William Kavanagh
% - [X] Marco Cook
% - [X] Craig Reilly (sp?)
% - [X] Zak Bevan
% - [X] Matt Barr
% - [X] Syed Waqar Nabi
% - [X] Derek Sommerville
% - [X] Quintin Cutts
% - [~] Everybody in the 510 lab
% - [X] Lovisa Sundin
% - [X] Patrizia LASTNAME
% - [X] Gözel Shakeri (sp?)
% - [~] RPGLite Players I can remember
% - [X] Frances LASTNAME
% - [X] PWSAfrica People
% - [~] Admin folk at Glasgow University (Helen, Gail)
% - [X] David Creigh (sp?)
% - [X] Leisha Hussein
% - [X] Emma Toussaint (check this is OK)
% - [X] Rita Toussaint
% - [X] Mati Araldo
% - [X] Francesco Porro
% - [ ] Luca Amato
% - [X] Sevki Harasci (sp?)
% - [X] Catherine Wallis
% - [X] Jamie Wallace
% - [X] AnnaSofia LASTNAME
% - [X] Rob Dekkers
% - [X] Mattia Salnitri
% - [X] George Campbell (Putney, Stall, Crisp)
% - [X] Milad Mehrabadi

}