
\begin{description}
  \item[actor —] allows the function to identify the actor performing the activity
    defined by the function. This argument is any object uniquely identifying an
    actor.
    \item[context —] allows the function to determine details of the current
      thread of work being undertaken by the actor. This is necessary because in
      some simulations, the same actor might pause and resume multiple
      occurrences of the same activity --- for example, they might concurrently
      play three different matches in RPGLite. As a result, it is necessary to
      understand the context of the action being performed by the actor in
      question. This argument can be any object uniquely identifying the context
      of a piece of work, but is most likely a class or dictionary-like object
      to permit the communication of information across invocations of different
      action-representing functions.
      \item[environment —] an actor's actions are often determined by the global
        environment they act within. There may be ancillary details to the
        actor's actions and the context of their particular thread of work which
        they are undertaken within which are used to determine behaviour, such
        as a landscape they traverse or other actors they might choose to
        interact with. Because all actors share access to a global environment,
        this also provides a message passing space, or a space where actors can
        set values and flags other actors might look to, should those details be
        more general than their specific thread of work at a given point in time.
\end{description}
