\chapter{Introduction}

\revnote{This introductory bit should be half its current length\ldots{}} This
thesis is about two distinct areas of study and what happens when they are
married together. First, it is concerned with \aop{}: a software engineering
paradigm which has received some academic interest, but comparatively little
industry use or practical demonstration of its benefits. It applies some extra
program logic (called ``advice'') to places where that program logic is required
(called ``join-points''); the combination of each is an ``aspect''. Integral to
the paradigm's design is its property of ``obliviousness'', where the
implementation of a join-point does not indicate whether an aspect applies to
it. Obliviousness simplifies parts of the codebase changed by aspects, as they
can be written without including special \aop{} logic. \Aspectorientation has
been well researched, but a specific context for which it is better suited than
its competitors is yet to be found. Its tooling and theory are well established,
but it isn't the ``tool of choice'' for any particular problem. Critics cite
obliviousness as a potential reason for this: reasoning about a program is
difficult if it could be altered by any other part of the program. However, a
use-case where \aop{} \emph{is} well-suited would allow the existing theory and
tooling to be put to use.

Second, it is concerned with how researchers write programs to support their
work. Many fields rely on software models --- programs which represent or
emulate a subject of study --- when conducting an investigation or experiment.
This software is subject to unique requirements, such as the ability to produce
reproducible results, to represent hypothetical systems or behaviours which may
not have been represented before, and an architecture supporting future studies
(potentially conducted by other research teams) which may look to extend that
software in ways which are hard to predict, as they encode novel ideas for
research. The design and maintenance of these codebases is unique as a result of
these requirements.

Later chapters motivate and demonstrate that \aop{} is well-suited to be used as
a tool for maintaining models, particularly in a research context. We suggest
that aspects can represent units of change to an existing model. This allows
research software engineers to develop experimental models as descriptions of
the differences between a system under study and a hypothesised one, by encoding
the differences between them as aspects. As later chapters discuss, the
properties which make the paradigm difficult to apply in industrial codebases
can be a strength in the context of research software engineering: as a model is
oblivious to the application of aspects, removing those aspects is trivial. One
can imagine that different models could be constructed by composing a ``base''
model with different sets of aspects. Failed experiments would not incur
technical debt, or necessitate the careful removal of code related to that
experiment.

\Aspectoriented{} modelling may be the application for which \aop{} is uniquely
well-suited, because obliviousness is a unique property of the paradigm.
However, aspects have not been used to date to represent units of change to
scientific models, and so the viability of the technique is unknown. This thesis
shows that \aspectoriented{} modelling is feasible by demonstrating that aspects
can be used to represent change to a modelled system. It investigates the use of
aspects to introduce new model parameters and actor behaviours, and shows that
aspects as units of change can be successfully applied to different systems in
certain cases.

This introductory chapter proposes the problem to be solved in research software
engineering more thoroughly in \cref{intro_problem_to_solve}, and proposes an
\aspectoriented{} solution in \cref{intro_proposed_solution}. The latter section
also describes \aop{} in more detail, and narrows the scope of the types of
models within the scope of this thesis. \Cref{intro_contributions} notes the
research questions and key contributions found in later chapters, and the
structure of the rest of the thesis is described in
\cref{intro_thesis_structure}.



\section{The Problem to Solve}
\label{intro_problem_to_solve}

Consider a research software engineer, responsible for the development and
maintenance of a well-adopted model of a system involving human behaviour, such as a
public transport system, the spread of a disease in a pandemic, or a software
development team's day-to-day work. As the model's use spreads and users'
requirements broaden, how might our engineer manage the maintenance burden of
their codebase? What tools and techniques might they make use of to ease their
task? Their tool might have originally modelled the interactions of individual
developers in a team (to extend the latter example), and offered parameters for
development methodologies or team sizes; subsequent researchers might look to
run simulations of the team with biased behaviour such as illness or tiredness,
or different experience levels in the team, or communication breakdowns between
individuals, or the impact of a change in management or project direction. What
then?

Typically, if a required change would alter the model significantly, researchers
looking to adopt the tool for new purposes would fork it (assuming the source is
available to them under a permissive enough licence) and would make whichever
modifications suited their needs. Other researchers might make their own forks.
If the change is small, or if the new behaviour can be enabled in configuration,
then it might be merged into the original tool. The contribution is thus
disseminated to the research community who adopted the original tool without the
need to migrate to a fork. However, if a future team wanted to research a
combination of factors --- say, simulating the impact of a change of management
on teams with different levels of experience and communication quality --- they
would repeat the process, producing another fork with their own implementation
of this particular combination. As modifications are built on top of each other,
the logic for each new behaviour must be interwoven with that of the original
model in any scenario where those changes cannot be developed as independent
modules for the original codebase.

Such a codebase would be increasingly difficult to maintain. For example, the
abstractions used to modularise it may have elegantly separated different
concerns into different modules at first, but abstractions are often
domain-specific. As the model becomes increasingly general-purpose, the
abstractions used to separate concerns accumulate technical debt in the face of
broader use-cases. In addition, the behaviours added by different teams may not
make sense when enabled at the same time. In the example of a model of a
software development team, a change representing remote work and a change
representing the spread of a virus in different office settings are at odds
conceptually and both are unlikely to be required in most cases. As a result,
the tool's source code risks becoming confusing and unwieldy.

There is therefore an opportunity to improve the design of software models used
in research contexts. We identify a need for a technique which allows a model to
be altered in an arbitrary fashion so that any changes required of the tool can
be represented as separate, optional modules --- even when that change
introduces or alters logic in the middle of an existing process, and therefore
would not modularise using other mechanisms. It should also permit a developer
to select the changes they need for their study, effectively composing a model
with whichever behaviours they require.



\section{A Proposed Solution}
\label{intro_proposed_solution}
We identify \aop{} as a tool which is well-suited to the needs of research
software engineers. This section starts with a definition of relevant terms and
an explanation of the scope of the research: the type of system being modelled,
``\sociotechnical{} systems'', are described in
\cref{intro_what_are_sociotechnical_systems}. This thesis concerns models and
simulations of \sociotechnical systems. Specific definitions of ``model'' and
``simulation'' for the purposes of this thesis are given in
\cref{intro_defining_simulation_and_model}.


Having defined the relevant terms, a high-level overview of \aop{} is given in
\cref{aop_brief_introduction}. Motivations for the construction of
\aspectoriented{} models can then be discussed, as the necessary background has
been given; this follows in \cref{intro_motivations_for_using_aop_in_models}. To
give the reader context as to the current research opportunities in the field of
\aop{} --- as well as the work required to build an \aspectorientation{}
framework suitable for \aspectoriented{} modelling --- some of \aop{}'s current
limitations are discussed in \cref{challenges_in_aop_briefly}. A more thorough
discussion of the aspect-oriented paradigm and its literature is given later in
\cref{aspect_oriented_programming_litreview}, and literature on its shortcomings
are also described in more detail in \cref{subsec:aop-criticisms}.



% We identify \aop{} as a tool which is well-suited to the needs of research
% software engineers. This section starts with a high-level introduction to
% aspect-oriented programming concepts in \cref{aop_brief_introduction}. Some
% use-cases for \aspectoriented{} modelling are given in
% \cref{intro_motivations_for_using_aop_in_models} to motivate the study.
% To give the reader context as to the current research opportunities in the field
% of \aop{} --- as well as the work required to build an \aspectorientation{}
% framework suitable for \aspectoriented{} modelling --- some of \aop{}'s current
% limitations are discussed in \cref{challenges_in_aop_briefly}.
% A more
% detailed discussion of the aspect-oriented paradigm and its literature is given
% later, in \cref{aspect_oriented_programming_litreview}, and literature on its
% shortcomings are also described in more detail in \cref{subsec:aop-criticisms}.

% The systems being modelled throughout this thesis are \sociotechnical systems.
% These concern interactions between human actors, technology, and business
% processes and are introduced in \cref{intro_what_are_sociotechnical_systems}. We
% expect this technique is broadly applicable to simulations and models generally.
% However, to limit the scope of the research undertaken in this thesis, the scope
% within which it is assessed is reduced, and it is examined as applied to
% \sociotechnical models in particular. These are also briefly introduced.

% Having introduced \aop{} and \sociotechnical systems, the
% section concludes by examining the technique's possible application in the
% context of the problem described.




\subsection{What are \SocioTechnical Systems?}
\label{intro_what_are_sociotechnical_systems}

The research in this thesis applies aspect-oriented programming to the modelling
\& simulation of \sociotechnical systems. A \sociotechnical system is one
composed of people, technology, and the interactions between the two. The
term originates in the study of work and organisations undertaken by
\citet{trist1951sociotechnical}. \Sociotechnical systems research continues to
focus on organisations and
workplaces~\cite{pasmore2019reflections,baxter2011socio}, though the interaction
of people and technology is also studied more broadly in areas such as degraded
modes~\cite{johnson2007degradedmodes}, resilience
engineering~\cite{hollnagel2006resilience}, and responsibility
modelling~\cite{lock2009responsibility}.

The involvement of human behaviour in the system's dynamics makes these systems
useful subjects for \aspectoriented{} modelling: variations in the behaviour of
an individual or an entire group is may be considered a concern which cuts
across different actors (or actors in models of different systems if the
behaviour is exhibited in more than one model). To this end, models of learning
are developed in \cref{chap:exp1_simulation_optimisation} which represent
learning as a cross-cutting concern. In addition, datasets of users'
interactions with computer systems can be collected to empirically verify
\sociotechnical{} models of their behaviour. This is also useful for the
research in later chapters: \cref{chap:rpglite} describes the design,
implementation, and data collected from the release of a mobile game, another
example of a \sociotechnical system.

The technique of creating models using \aspectorientation to modularise
particular behaviours or dynamics is not uniquely useful in the domain of
\sociotechnical systems modelling.
We suspect that it can be generalised to the behaviours and dynamics
of other systems, too. However, the paradigm is
demonstrated as a \sociotechnical modelling technique for the purposes of
setting bounds on our aims. It is not feasible to verify the
technique's appropriateness in \emph{every} type of system, but it is feasible to
investigate the technique as applied to a \emph{particular} type of system.
Also, aspect-orientation has received attention in the business process
modelling research
community~\cite{charfi2007ao4bpel,Cappelli_AOBPM,Charfi2006AspectOrientedWL},
setting a precedent for its broader use in modelling other \sociotechnical
systems too.


\subsection{Simulations \& Models}
\label{intro_defining_simulation_and_model}

The field of simulation \& modelling concerns the building of models of some
system and the simulation of that system using the model. We have found
(anecdotally) that the difference between the two can be vague; some definitions
are provided to avoid confusion.

In this thesis,``model'' will be used to refer to some representation of a
system or subject of study, or an abstraction of it. In this sense, a model can
be a concept, a physical model, a diagram, and so on --- in this thesis in
particular, ``model'' will be used with the meaning: \emph{``a software
representation of a system or other subject of study''}. An example of such
models are OBASHI models~\cite{obashimethodology}, which are software
representations of how data flows through \sociotechnical systems. Another is
CovidSIM~\cite{covidsim_tool}, which models the spread of COVID-19 through a
population by representing proportions of that population at different points of
a pandemic who are susceptible to infection, exposed to the virus, presently
infected, or recovered from infection.

The term ``simulation'' also requires definition, and throughout this thesis
will generally refer to the execution of a model. More formally put,
``simulation'' is used to refer to \emph{``the emulation of the processes within
a system or other subject of study''}. A simulation typically generates data,
such as a record of system states over time or a log of actions taken by some
actor within that system. One example of such a simulation could be the
execution of a model of stock trading where agents mimic the activity of
real-world traders to study effective trading behaviour~\cite{ZipTradersCliff}.
The software artifact represents the process of trading; executing that process
mimics trading itself, and so simulates the activity by emulating real-world
behaviours.


As the research in later chapters is primarily concerned with the simulation \&
modelling of \sociotechnical systems, the subject of a simulation or model
should be assumed to be a \sociotechnical system unless otherwise specified.
When models are discussed in the context of the research conducted in later
chapters, they refer specifically to models constructed for the purposes of
simulation. Models designed for purposes other than simulation may also prove to
be suitable for the use of \aspectoriented{} modelling, such as non-executable
models of business processes or
dataflow~\cite{obashimethodology,Cappelli_AOBPM,charfi2010AO4BPMN}. However,
models designed for simulation purposes are the focus of later chapters to limit
the scope of the research at hand.




\subsection{What is \AOP{}?}
\label{aop_brief_introduction}

\Aop{} originated in \citeauthor{kiczales1997aspect}'s work
at Xerox Parc~\cite{kiczales1997aspect}. The motivation for the paradigm was the
modularisation of ``cross-cutting concerns'': parts of a program which are not
directly related to the fulfilment of its requirements but are common throughout
it. Cross-cutting concerns generally don't relate to their neighbouring logic
semantically. Rather, they fulfil an ancillary task, such as logging. This
suggests it is a suitable candidate for modularisation: the single
responsibility principle~\cite{martin2003singleresponsibility} suggests that
software with multiple concerns should be refactored into separate modules.
However, traditional techniques are unsuitable for the creation of separate
modules in these situations.

Aspect-oriented programming introduces a new pattern which modularises
cross-cutting concerns. This is achieved by separating them from a program's
core logic and rewriting them as modules (``advice''), the combination of some
implementation (an ``aspect'') and places in a program to apply it
(``join-points'', a set of which are defined by a ``pointcut'') --- which are
later ``woven'' back into their intended positions by ensuring the aspect's
logic is included somehow at its corresponding join point. This allows
high-level behaviours such as logging, synchronisation, or memory safety to be
factored away from a program's logic and into separate units. Unlike traditional
modules, which are included by the code which requires them, a ``weaver''
composes advice and the code they should be applied to, with neither requiring
knowledge of the other. This separation is core to aspect-oriented programming's
design, and is referred to as ``obliviousness''.

Aspect-orientation allows for existing codebases to be structured in new ways,
but its notion of modularising cross-cutting concerns is general and can be
applied in other fields too. For example, business process modelling languages
have been adapted to accommodate behaviours or processes modelled using
advice~\cite{Cappelli_AOBPM,da2020implementation,charfi2007ao4bpel}. Some
researchers suggest using aspect-oriented programming to separate concepts in
experimental software, such as experiment and observation, to mirror the setup
of a traditional experiment in research-specific codebases~\cite{gulyas1999use}.
Concepts originating in \aop{} have also seen use outside of software
engineering~\cite{Cieslak_2011,Cappelli_AOBPM,da2020implementation,charfi2007ao4bpel}.

\revnote{
  This can be reworked to refer to variance less in line with the approach to
  discussing the research I now take in \cref{chap:experiment_setup} and
  \cref{chap:experimental_results}. However it's not essential, so I'll revisit
  \emph{if time permits}.
}This thesis explores modelling changes to behaviour using advice, and is applied
to the modelling of \sociotechnical systems in particular. \Sociotechnical
systems partly consist of the behaviour of human agents within a system, which
can exhibit variations contingent on environmental factors and other states.
This variance might be expressed differently in different agents in the system.
In later chapters, we demonstrate that these behavioural variations can be
separated from a generic model into advice, with the advantage that the generic
model's codebase is simplified and any particular behaviour of interest in a
particular study can be modelled by weaving those behaviours at runtime. Advice,
as demonstrated in later chapters, is a flexible and useful tool for composing
models.

\subsection{Motivations for \AspectOriented{} Modelling}
\label{intro_motivations_for_using_aop_in_models}

A researcher who maintains a model of some system they are studying would have
many potential use-cases for \aop{}. Some are discussed here to motivate the use
of \aop{} in a research software engineering context.

First, a researcher looking to study new behaviours for actors in their model
would need to implement those behaviours: however, the original behaviour of the
model may be useful as a control or point of comparison in their study. They may
``fork'' their codebase to produce two models (one unmodified, and one with
their new behaviour), but this requires them to maintain another codebase every
time a new behaviour is studied. A single codebase containing behavioural
changes as optional modules would reduce the overhead of maintaining that
codebase: changes to the original model could be made once, instead of in every
fork.

Some changes do not modularise using existing
techniques~\cite{kiczales1997aspect} and opportunities to do so depend on
codebase being changed. For example, an epidemiological model which represents
interactions between people in a bar would yield different possible changes
depending on the implementation of the bar's physical layout, distances between
people, places where they may interact, and the nature of those interactions.
The software itself would yield different opportunities to implement changes as
modules depending on the programming language, paradigm, and abstractions used.

However, \emph{modifications represented as aspects can be applied independently
of these factors}. This is a consequence of \aspectorientation{}'s unique
property of obliviousness, which is that aspects can be applied to an existing
codebase regardless of its implementation, because it does not need to include
any special structure or logic for aspects to be applied\cite{filman2000aspect}.
This means that the epidemiological model mentioned earlier could represent new
types of interactions between modelled agents regardless of whether the codebase
was designed to easily support new kinds of interactions, by encoding them as
aspects.

A second scenario where researchers may look to represent model changes as
advice would be in the instrumentation of that model for experimental
observation~\cite{gulyas1999use}. For example: researchers may develop a model
of a city's transport infrastructure to measure commute times which outputs
measurements of simulated commuters' journey duration. Other researchers --- or
future projects conducted by the same research team --- may look to extend that
model to also measure the \emph{expense} of those commutes, to contrast the time
and money spent commuting from different suburbs. While simulated commuters'
behaviour may not require modification, the measurements made of their commutes
would. These observations can be made by weaving aspects which record additional
information about modelled commutes without requiring modification of the
original codebase, which is potentially onerous (as described in this
subsection). This design for an experiment implemented as software mirrors
experiment best practices in other fields, where observational apparatus is
separated from a subject under study so as not to bias
results~\cite{gulyas1999use}.

A third scenario where \aspectorientation{} is useful relates to the second:
instrumenting several models in the same way to study how those models compare.
\Aop{}'s obliviousness allows researchers to apply advice to \emph{any}
codebase, including those which were not developed with \aspectorientation{} in
mind. For example, comparisons of multiple models could be performed in the same
way by applying the same aspects to each codebase to observe their states during
execution. Previous studies which compared software models of physical phenomena
such as the formation of galaxies~\cite{comparison_of_galaxy_formation_models}
or the radiation emitted by
them~\cite{comparison_of_radiative_models_of_galaxies} contrast the outputs of
those models but could not observe their state during execution. More detailed
measurements could be made by instrumenting each model for observation during
their execution. This could be achieved by applying aspects to model codebases
to take the same measurements of every model being compared regardless of their
architecture, taking advantage of \aop{}'s property of obliviousness.


Another motivating use-case for \aspectoriented{} modelling is that researchers
could use \aop{} to design experiments. For example, consider a study which
seeks to compare the impact of several political policies on a country's
economy, environment, and health to find a set of policies which optimise them.
Policies could be represented as changes to a model of a country as it presently
exists, such as World3~\cite{meadows1972limits}.\correctionnote{Include a
diagram of the model being proposed. Consider including terms like ``base
model'' in the diagram, introducing those terms here, and using them to make
this example clearer.} By implementing $n$ policies as aspects, up to $2^{n}$
models can be produced by applying every combination of aspects to the country's
model. The optimal set of policies can be discovered by comparing the outlook of
the country with each combination of policies applied. This has several benefits
over a study where a model is developed with policies included directly in the
model codebase. For example, the study can be applied to other countries by
changing the underlying country without risk of compromising the implementation
of any policy, as their implementations are independent. Extending the study is
also simple: adding policies to the study would only require writing new
aspects, rather than maintaining a potentially large codebase. This also avoids
the risk of introducing bugs to the implementations of other policies, as well
as avoiding the introduction of bugs to the model of the country itself. 

These motivating cases are all made possible due to \aop{}'s unique property of
obliviousness~\cite{filman2000aspect}. These results therefore cannot be
achieved using other modularisation techniques. Researchers who build, maintain,
or modify models may have a need for \aspectoriented{} modelling in any of the
scenarios described, and \aop{} may be a useful tool for research software
engineers.



\subsection{Challenges in \AspectOrientation{} Today}
\label{challenges_in_aop_briefly}

The \aspectoriented{} paradigm has some drawbacks. Identifying these means they
can be addressed when applying \aop{} to a new domain, and add context which
motivates research presented in later chapters. While they are discussed in
detail in \cref{subsec:aop-criticisms}, a quick summary is given here as
background for the ease of reading later chapters and for the discussion of the
contributions made in this thesis, which follows in \cref{intro_contributions}.

The paradigm's philosophy of obliviousness can make codebases difficult to
comprehend. When reading the implementation of a function or feature, it's not
possible to know whether advice is being applied to it elsewhere by inspecting
the relevant source code in isolation. If aspects are applied to it, the
behaviour of that code might be altered. Accurately understanding a program's
control flow can therefore be challenging as a consequence of aspect-oriented
programming's fundamental design
properties~\cite{Constantinides04aopconsidered,przybylek2010wrong}.

Other limitations of aspect-oriented programming concern aspects themselves.
\Aspectorientation{} tooling typically allows for advice to be woven before
join-points, after them, or around them (in effect, both before and after).
However, this requires aspects to treat their join-points as black boxes: no
changes can be made \emph{inside} the join-point. This could prevent some
use-cases of \aspectoriented{} modelling, such as the instrumentation of several
models to compare them (discussed in
\cref{intro_motivations_for_using_aop_in_models}). This would only be feasible
if the join-points available in a codebase allowed aspects to collect the
information required. If those join-points are treated as black boxes and the
state to observe only exists inside them, the instrumentation suggested would
not be possible. Therefore, \aspectoriented{} modelling frameworks would ideally
support new kinds of join-point which advice can be applied inside.

Lastly, while a variety of toolsets exist for \aop{}, few studies appear to have
been undertaken confirming that its expected benefits actually
materialise~\cite{przybylek2018empirical}. Some scepticism around its practical
benefits have been
raised~\cite{steimann06paradoxical,przybylek2010wrong,Constantinides04aopconsidered}.
An application of aspect orientation to a new domain ought to demonstrate that
the technique is of practical benefit, rather than a theoretical curiosity.



\correctionnote{I've commented out a subsection below which explained some
possible benefits of \aspectoriented{} modelling, because I've written a new
``motivations for \aspectoriented{} modelling'' section above. Consider what to
do with these two…}

% \subsection{Possible Benefits of Aspect-Oriented Modelling}
% Some specific benefits we anticipate from an aspect-oriented approach to
% building and maintaining models include:

% \Needspace{3\baselineskip}
% \begin{description}
%   \item[Advice as Units of Model Change] Scientific models which already exist
%     are difficult to modify. They can be adopted by many research groups,
%     meaning breaking changes impact the broader community; they can be the basis
%     of published results, so changing the source code might invalidate the
%     relationship between ongoing work and published work; and they can be
%     brittle, as the incentive when writing software for research purposes is to
%     achieve results worth publishing, rather than to produce a high-quality and
%     maintainable codebase. This differs from a commercial software engineering
%     team's incentives, which are typically to produce software which they can
%     continue to produce in the future with minimal overhead imposed by code
%     quality. For these reasons, scientific codebases have special requirements
%     which discourage direct modification, particularly for different use-cases.
%     Aspect-oriented programming allows updates to codebases to be written
%     without direct modification to the source code. In the case of research
%     software specifically, this has been shown to achieve positive results in
%     previous case studies~\cite{ionescu2009aspect,wallis2018caise}.

%   \item[Advice as Tools for Instrumenting Scientific Codebases]
%     Models of a system are ideally concerned with the logic required to
%     accurately model the system itself, and do not contain additional logic to
%     instrument the model for the purposes of a particular experiment. This is
%     desirable because it allows the model to be re-used for many experiments, as
%     the instrumentation to make observations for a particular purpose are not
%     woven throughout its logic. This also makes the codebase easier to read: a
%     researcher interested in it only needs to read the logic which was required
%     to implement a model, and doesn't also have to identify the parts of the
%     program which don't model a system at all (but observe a simulation of it
%     instead). Finally, the separation of observational apparatus from a model
%     mirrors the design of a traditional experiment, where observations are
%     carefully made so as not to bias results. This potential use-case of aspect
%     orientation has been suggested in the community~\cite{gulyas1999use}.

%   \item[Advice as Hypothesised System Behaviour] If advice is woven into a
%   model, side effects of the advice could be used to alter the model. In this
%   scenario, advice could be written which changes a model to have a desired
%   effect, such as correcting an error in the model or updating it in light of
%   new research. Moreover, an established model may serve as a ``proving ground''
%   for future research: hypothesised behaviours in a system may be written as
%   updates to a proven model of that system, and its predictive quality compared
%   to the model executed without advice woven. A simple experimental structure is
%   yielded by this process. Given advice may be useful as a unit of model change,
%   these units may be proposed, tested, and combined to produce a variety of
%   models to meet different research groups' needs without requiring onerous
%   maintenance efforts.
% \end{description}




\section{Contributions}
\label{intro_contributions}

\subsection{Research Questions}
\label{intro_rqs}

The research questions investigated in the following chapters are: 

\begin{researchquestion}
\begin{description}
\item[RQ1] \rqone{}
\item[RQ2] \rqtwo{}
\item[RQ3] \rqthree{}
\end{description}
\end{researchquestion}

These are motivated following a review of relevant literature in
\cref{subsec:rqs}. Clarifications of specific terms and phrases are also
provided in \cref{subsec:rqs} as it contains a thorough discussion
of the research questions.

\subsection{Summary of Contributions}

In developing tooling for aspect-oriented modelling and investigating the
technique's feasibility, this thesis makes multiple contributions to the
relevant research communities, including but not limited to answering the above
research questions. These are summarised as:

\Needspace{3\baselineskip}
\begin{description}
  \item[Tooling for \AspectOriented{} Modelling] This thesis presents a
    redesigned and re-implemented version of a tool, \pydysofu{}\footnote{\pydysofu{} is
    an acronym for \emph{PYthon DYnamic SOurce FUzzer}.}, originally prototyped
    for some prior work,\footnote{See \cref{chap:prior_work} for a discussion on
    related research undertaken prior to starting this PhD.} which contributes
    both a tool which can be employed for aspect-oriented modelling and a
    weaving technique designed for legibility when applying aspects to models.
    It can be applied to existing Python code without modifying the codebase it
    is applied to and introduces minimal dependencies when added to a project.
  \item[Dataset describing \SocioTechnical System Interaction] A dataset
    describing 370 players' interactions with a mobile game released for
    research purposes called RPGLite. The design, implementation, and data
    obtained from the game is presented, and the dataset collected via RPGLite
    is used to design experiments presented in this thesis.
  \item[Demonstration of Aspect-Oriented Model Enhancement] Aspects are used to
    augment a model of RPGLite gameplay and demonstrate that aspect-oriented
    enhancements can alter a model to improve it. This is achieved by measuring
    the similarity of simulated players' preference of characters in the game
    with that of real-world players by correlating the frequency with which
    simulated and real players select different characters. In particular, we
    show that a model can be augmented to synthesise data with the properties of
    the real-world dataset.
  \item[Using Aspect-Oriented Enhancements to Identify Hypothesised Player
    Behaviour] A model is presented which represents how players learn which
    characters are likely to win games of RPGLite. We demonstrate that this can
    be tuned to specific real-world players' behaviour using the dataset
    collected, resulting in player-specific models of learning which characters
    are likely to win games. Properties of learning of specific real-world
    players are identified by fitting a learning model to their RPGLite gameplay
    data.
  \item[Investigation into Aspect Portability] We investigate whether
    aspect-oriented changes to a model can be ported from one system to another,
    taking advantage of aspect-orientation's modular nature.
  \item[Exploration of Opportunities Enabled by Aspect-Oriented Modelling]
    New research opportunities are yielded by this novel technique in simulation
    \& modelling. As in other PhD theses where contributions carry potential for
    many pieces of future work~\cite{marsh1994formalising},\correctionnote{Phil
    questions whether this is an appropriate citation for claiming this as a
    contribution --- I'm not sure how to change this. If there wasn't a citation
    he'd want one...and I'm basing the claim off Steve's thesis specifically.
    Considering leaving this as-is. Revisit before submitting corrections.} the avenues for
    future research are broad enough that identifying and enumerating them is a
    substantial piece of work and constitutes an additional contribution.
\end{description}


% \section{Terms \& definitions}\label{sec:glossary}

% \inline{Complete the glossary in \cref{sec:glossary}.}

% \inline{Decide whether terms like BPMN, simulation \& modelling, etc also belong
% in the glossary\cref{sec:glossary}.}

% \inline{Dejice's thesis has a similar glossary; how did he structure it? Wasn't
%   an appendix. Was it a section of the intro or something else?}

% \begin{description}
%   \item[\AOP{}]
%   \item[Cross-Cutting Concern]
%     \item[Aspect] 
%     \item[Join Point]
%     \item[Advice] 
%     \item[Point Cut] 
%     \item[Weaving] 
%     % \item[AspectJ] The original aspect orientation framework, with language
%     % extensions to describe pointcuts and aspects.
%     \item[Target] The procedure an aspect is applied to via a join point, to
%     affect advice.
%     \item[\pdsf{}] 
%     \item[]
% \end{description}



\section{Thesis Structure}
\label{intro_thesis_structure}

The rest of the thesis is structured as follows.

\Cref{chap:lit_review} surveys the project's relevant literature and identifies
specific research questions in the field which the thesis addresses. Some
earlier work precedes the research presented in this thesis. To delineate this
from the contributions presented in later chapters, \cref{chap:prior_work}
surveys the state of the research project before this PhD began.

Having surveyed the literature, identified research questions, and established
the starting point of the research, \cref{chap:pdsf_rewrite} describes the
re-design and implementation of the \aop{} framework \pydysofu{}, named
\pdsfthree. Other technical contributions follow in \cref{chap:rpglite}, which
describes the design and implementation of RPGLite, a mobile game developed for
research purposes, as well as the data collected from it.

Later chapters explore the application of \aop{} to simulation \& modelling
codebases. Three experiments are constructed using a model of RPGLite to answer
the research questions mentioned in \cref{intro_rqs}. The experiments involved
share the majority of their technical foundations and methodology, but are used
in different ways to research different facets of \aop{} as applied to
simulation \& modelling. For this reason, their common foundations and
methodology are explained in \cref{chap:experiment_setup}. The specifics of each
experiment --- and the results of those experiments --- are described in their
own chapter, \cref{chap:experimental_results}, alongside answers to the research
questions they address.

Finally, the encoding of behaviours and model properties as aspects yields novel
research approaches, a discussion of which is omitted in all relevant literature
reviewed to date. We therefore investigate the possibilities enabled by
\aspectoriented{} modelling in \cref{chap:future_work} thoroughly, in the same
vein as other theses making similar contributions~\cite{marsh1994formalising}.







