\chapter{Complete Description of Model Parameters}
\label{appendix_model_parameters}

The \lstinline{ModelParameters} class defines the parameters which generate data
from an experiment. It is introduced in \cref{model_parameters}, and its
implementation is shown in \cref{fig:model_parameters_class} and
\cref{fig:ModelParameters_run_experiment_method}. However, many parameters
control the 

  

\begin{description}[style=multiline,leftmargin=3cm]
  \item[starting\\confidence] The initial value of the model's confidence curve.
  \item[assumed\\confidence\\plateau] A corresponding high value for the model's
  confidence curve. As the curve's growth rate is calculated to align with the
  number of games completed by the real-world player being simulated, and the
  curve asymptotically approaches 1, a high value is required which represents
  the expected confidence of a real-world player having played a significant
  number of games.
  \item[curve\\inflection\\relative to\\numgames] The proportion of games played by
  the real-world player being simulated at which the player's confidence
  curve should reach the parameter \lstinline{assumed_confidence_plateau}. This
  allows for control over the growth rate of the curve while linking the rate of
  growth to the number of games played by the real-world player being simulated;
  \item[C] The curve parameter of the confidence model's underlying birch curve;
  \item[prob\\bored] The probability that a player with confidence above
  \lstinline{assumed_confidence_plateau} becomes ``bored'' and stops playing
  RPGLite. These players are removed from the simulated playerbase and replaced
  with new players. This mechanism is discussed in further detail in
  \cref{subsec:controlling_state_space_exploration}.
  \item[boredom\\enabled] Whether the boredom mechanism discussed in
  \cref{subsec:controlling_state_space_exploration} is to be enabled.
  \item[training\\data] The set of real-world games used as a training fold when
  optimising parameters for simulating a player. This is used when annealing
  toward optimal parameters for the learning model, and so is explained in more
  detail in \cref{methodology_explained}.
  \item[testing\\data] The set of real-world games used as a testing fold when
  optimising parameters for simulating a player. This is used when annealing
  toward optimal parameters for the learning model, and so is explained in more
  detail in \cref{methodology_explained}.
  \item[iteration\\base] The number of games to play when generating a synthetic
  dataset. If \lstinline{boredom_enabled==False}, the number of games is instead
  calculated to ensure that all synthetic players play the same number of games
  as were completed by the played being simulated, (calculated from the training
  or testing fold as appropriate).
  \item[number\\simulated\\players] The size of the simulated playerbase.
  \item[advice] A list of tuples defining advice to weave when generating data.
  Pieces of advice are uniquely defined by a tuple containing the type of aspect
  to weave (such as \lstinline{"prelude"} or \lstinline{"error_handler"}), a
  string-represented regular expression defining the join point of the advice,
  and a \lstinline{callable} to use as an aspect when weaving the advice. As
  \lstinline{ModelParameters} instances are serialised to disk to persist
  experiment results and Python functions are not seralisable, this may also be
  a string value; strings are IDs of aspects, and are replaced with their
  corresponding aspect on deserialisation.
  \item[players] The player being simulated. This parameter is a list of player
  usernames to support simulating groups of players if required, though this
  functionality is not used in the experiments presented in
  \cref{chap:experimental_results}.
  \item[args] Any additional arguments to pass when running a simulation.
  \item[kwargs] Any additional keyword arguments to pass when running a simulation.
  \item[boredom\\period] The number of games completed by the simulated
  playerbase between checks of players' confidence and random selection for
  removal (default 25).
  \item[initial\\exploration] A value which controls the number of
  games individual players complete before their confidence \& learning models control
  their character pair selection (default 28, to allow sufficient games to
  select every available pair exactly once).
\end{description}
  
